Geometric models reveal behavioral and neural signatures of transforming naturalistic experiences into episodic memories

### 2020-09-02 paper/main.tex
recompiled pdf with updated figs
center fig 5
updated title

\title{Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories}

@@

\section*{Introduction}
What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete binary operation: each studied item may be separated from the rest of one's experience and labeled as having been either recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between recollecting the (contextual) details of an experience or having a general feeling of familiarity~\citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of information regarding human episodic memory~\citep[for review see][]{Kaha12}.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep[for review, also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a naturalistic event from the context in which it occurs can substantially change its meaning.  Second, whether or not the rememberer has precisely reproduced a specific set of words in describing a given experience is nearly orthogonal to how well they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of \textit{exact} recalls is often considered to be a primary metric for assessing the quality of participants' memories.  Third, one might remember the essence (or a general summary) of an experience but forget (or neglect to recount) particular low-level details.  Capturing the essence of what happened is often a main goal of recounting an episodic memory to a listener, whereas the inclusion of specific low-level details is often less pertinent.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_detail}
  \caption{\small \textbf{Precision reflects the completeness of recall, whereas distinctiveness reflects recall specificity.} \textbf{A.} Recall precision by episode event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single episode event.  Colored dots within each violin plot represent individual participants' recall precision for the given event.  episode events are ordered along the $x$-axis by the average precision with which they were remembered.  \textbf{B.} The set of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising an example episode event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{C.} Excerpts from the most precise (P17) and least precise (P6) participants' recalls of episode event 22.  Descriptions of specific actions or features reflecting those highlighted in Panel B are highlighted in the corresponding color.  The text highlighted in gray denotes a (rare) false recall.  The unhighlighted text denotes correctly recalled information about other episode events.  \textbf{D.} Recall distinctiveness by episode event.  Kernel density estimates for each episode event's distribution of recall distinctiveness scores, analogous to Panel A.  \textbf{E.} The sets of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising episode events described by the example participants in Panel F.  Each event's text is highlighted in a different color.  \textbf{F.} The sentences comprising the most distinctive (P9) and least distinctive (P6) participants' recalls of episode event 22.  Sections of recall describing each each episode event in Panel E are highlighted with the corresponding color.}
  \label{fig:precision-detail}
\end{figure}

>>>>

\title{Geometric models reveal behavioral and neural signatures of transforming naturalistic experiences into episodic memories}

@@

\section*{Introduction}
What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete binary operation: each studied item may be separated from the rest of one's experience and labeled as having been either recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between recollecting the (contextual) details of an experience and having a general feeling of familiarity~\citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of information regarding human episodic memory~\citep[for review see][]{Kaha12}.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep[for review, also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a naturalistic event from the context in which it occurs can substantially change its meaning.  Second, whether or not the rememberer has precisely reproduced a specific set of words in describing a given experience is nearly orthogonal to how well they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of \textit{exact} recalls is often considered to be a primary metric for assessing the quality of participants' memories.  Third, one might remember the essence (or a general summary) of an experience but forget (or neglect to recount) particular low-level details.  Capturing the essence of what happened is often a main goal of recounting an episodic memory to a listener, whereas the inclusion of specific low-level details is often less pertinent.

@@

\begin{figure}[tp]
  \centering
  \vspace*{-1cm}
  \includegraphics[width=.8\textwidth]{figs/precision_distinctiveness_detail}
  \caption{\small \textbf{Precision reflects the completeness of recall, whereas distinctiveness reflects recall specificity.} \textbf{A.} Recall precision by episode event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single episode event.  Colored dots within each violin plot represent individual participants' recall precision for the given event.  episode events are ordered along the $x$-axis by the average precision with which they were remembered.  \textbf{B.} The set of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising an example episode event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{C.} Excerpts from the most precise (P17) and least precise (P6) participants' recalls of episode event 22.  Descriptions of specific actions or features reflecting those highlighted in Panel B are highlighted in the corresponding color.  The text highlighted in gray denotes a (rare) false recall.  The unhighlighted text denotes correctly recalled information about other episode events.  \textbf{D.} Recall distinctiveness by episode event.  Kernel density estimates for each episode event's distribution of recall distinctiveness scores, analogous to Panel A.  \textbf{E.} The sets of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising episode events described by the example participants in Panel F.  Each event's text is highlighted in a different color.  \textbf{F.} The sentences comprising the most distinctive (P9) and least distinctive (P6) participants' recalls of episode event 22.  Sections of recall describing each each episode event in Panel E are highlighted with the corresponding color.}
  \label{fig:precision-detail}
\end{figure}
